---
title: "Roulette Simulation"
author: "Sarah Torrence"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document: 
    code_folding: hide
    toc: yes
    number_sections: true
    toc_depth: 3
    toc_float: true
---

```{r global options, include = FALSE}
knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
library(dplyr)
```

# Introduction

If you have any interest in gambling you would be familiar with the game Roulette. Further, if you have played Roulette you might be familiar with the classic strategy called the Martingale strategy. Using some simple math and logic it seems that the Martingale strategy leaves the player with positive earnings, but is that really true? We know gambling games are typically set up so the house wins in the long run making the Martingale strategy seem too good to be true. Using simulation, I have set out to test the Martingale strategy to determine whether it really works. 

## Background

<img src="./assets/roulette-table.svg" style="width:50.0%" />

In the American version of Roulette, the wheel consists of 38 evenly sized pockets 18 of which are red, 18 black and 2 green as shown in the above image. These pockets are also numbered up to 36 including 0 and 00. In the game, a pocket is randomly selected. Before selection, gamblers may wage on the outcome in several different ways including selecting a specific number or color trying to predict the outcome.

For our purposes, we will ignore the numbers and only use the colors of red, black and green. For each $1 wager a gambler places on a specific color, they will win $1 if the color chosen is the outcome of the game. For example, if the gambler bets $1 on red and a red pocket is in fact randomly selected, the gambler would be given $2 (the original $1 wager and the additional $1 of winnings). If red is not selected, the gambler would be at a loss of $1 which was their original wager. 

Using the Martingale strategy, a gambler would chose a color, red or black, and stick with this color for each round. Their starting wager would be $1. If the outcome is not the color of their choice, they would double their wager to $2 and continue doubling their wager until they are successful. The round after their chosen color is selected they will go back to their original wager of $1.

![](./assets/martingale-strategy.svg)

Consider the below sequence of spins with the following outcomes in which the gambler chose to wager money on Red as the outcome.

| Play | Wager | Outcome | Earnings |
|:----:|:-----:|:-------:|:--------:|
|   1  |   1   |  Black  |    -1    |
|   2  |   2   |  Black  |    -3    |
|   3  |   4   |   Red   |    +1    |

As you can see they lost their wager the first round, bet $2 the second round, lost that money as well, bet $3 the third round and were successful returning an earnings of $1 after the 3 spins.Now let's consider another sequence: Black, Black, Black, Red.

| Play | Wager | Outcome | Earnings |
|:----:|:-----:|:-------:|:--------:|
|   1  |   1   |  Black  |    -1    |
|   2  |   2   |  Black  |    -3    |
|   3  |   4   |  Black  |    -7    |
|   4  |   8   |   Red   |    +1    |

It appears that using the Martingale strategy, the gambler always comes out ahead with earnings of $1. Does this still hold true over many iterations? 

# Methods

Here I will simulate the Martingale strategy to find the average earnings of a gambler assuming the gambler wagers on the spin selecting a red pocket with a probability of 18/38. The parameters are the budget (amount of money the gambler can spend) (**B**), the budget threshold for successfully stopping (**W**), the maximum number of plays allowed (**L**) and the casino's wager limit (**M**). Using these parameters, I conduct a series of single spins in which the previous result (win/loss) is used to determine whether the wager in the next spin is $1 (if the previous was spin was red) or double the previous wager (if the spin was black or green). This series of spins continues until the gambler runs out of money (he/she has none of their original budget remaining), hits the maximum number of plays, or reaches the threshold for successfully stopping.

Below is a table showing the parameters used and their starting values for purposes of this simulaion:

| Parameter | Description                     |             Starting value             |
|:---------:|:--------------------------------|:--------------------------------------:|
|   **B**   | Starting budget                 |                  $200                  |
|   **W**   | Winnings threshold for stopping | $300 (Starting budget + $100 winnings) |
|   **L**   | Time threshold for stopping     |               1000 plays               |
|   **M**   | Casino’s maximum wager          |                  $100                  |


```{r}
#' A single play of the Martingale strategy
#'
#' Takes a state list, spins the roulette wheel, returns the state list with updated values (for example, budget, plays, etc)
#' @param state A list with the following entries: 
#'   B              number, the budget
#'   W              number, the budget threshold for successfully stopping
#'   L              number, the maximum number of plays 
#'   M              number, the casino wager limit
#'   plays          integer, the number of plays executed
#'   previous_wager number, the wager in the previous play (0 at first play)
#'   previous_win   TRUE/FALSE, indicator if the previous play was a win (TRUE at first play)
#' @return The updated state list
one_play <- function(state){
  
    # Wager
    proposed_wager <- ifelse(state$previous_win, 1, 2*state$previous_wager)
    wager <- min(proposed_wager, state$M, state$B)
    
    # Spin of the wheel
    red <- rbinom(1,1,18/38)
    
    # Update state
    state$plays <- state$plays + 1
    state$previous_wager <- wager
    if(red){
      # WIN
      state$B <- state$B + wager
      state$previous_win <- TRUE
    }else{
      # LOSE
      state$B <- state$B - wager
      state$previous_win <- FALSE
    }
  state
}


#' Stopping rule
#'
#' Takes the state list and determines if the gambler has to stop
#' @param state A list.  See one_play
#' @return TRUE/FALSE
stop_play <- function(state){
  if(state$B <= 0) return(TRUE)
  if(state$plays >= state$L) return(TRUE)
  if(state$B >= state$W) return(TRUE)
  FALSE
}


#' Play roulette to either bankruptcy, success, or play limits
#'
#' @param B number, the starting budget
#' @param W number, the budget threshold for successfully stopping
#' @param L number, the maximum number of plays 
#' @param M number, the casino wager limit
#' @return A vector of budget values calculated after each play.
one_series <- function(
    B = 200
  , W = 300
  , L = 1000
  , M = 100
){

  # initial state
  state <- list(
    B = B
  , W = W
  , L = L
  , M = M
  , plays = 0
  , previous_wager = 0
  , previous_win = TRUE
  )
  
  # vector to store budget over series of plays
  budget <- rep(NA, L)
  
 # For loop of plays
  for(i in 1:L){
    new_state <- state %>% one_play
    budget[i] <- new_state$B
    if(new_state %>% stop_play){
      num_plays <- new_state$plays
      return(data.frame(budget[1:i],num_plays))
    }
    state <- new_state
  }
  budget    
}

# helper function
get_last <- function(x) tail(x,n=1)
```

I ran the simulation once to replicate the process of one player using this strategy with the above starting values and graphed the relationship between the gamblers earnings and the number of rounds of Roulette they had played.

```{r, cache=TRUE}
#Simulating one play
first_play <- one_series(B = 200, W = 300, L = 1000, M = 100)

#Plotting the earnings based on play number
plot(x=c(1:length(first_play[,1])),y=first_play[,1]-200, type="l",xlab = "Play Number", ylab = "Earnings",main="One Player Simulation")
```

Here is the graph of a second player's simulation again with the same starting values as above.

```{r, cache=TRUE}
#Simulating a second play
second_play <- one_series(B = 200, W = 300, L = 1000, M = 100)

#Plotting the earnings based on play number
plot(x=c(1:length(second_play[,1])),y=second_play[,1]-200, type="l",xlab = "Play Number", ylab = "Earnings",main="A Second Player Simulation")
```

You can see from both of these graphs that each player started out with a gradual increase in earnings followed by several large troughs where they lost quite a bit of money. In each scenario the gambler's never had more than $50 in earnings and were in the negative earnings, meaning they were losing money for most of their rounds. Of course this is only 2 scenarios, but we want to know what kind of odds a gambler has of making money from this strategy overtime. Below we will run the simulation many more times to get a better estimated average earnings.

# Results

To determine the outcome of average earnings using the Martingale strategy I ran the simulation 10,000 times with the starting budget of $200, the budget threshold of $300, the maximum number of plays of 1,000 and a casino wager limit of $100.

In each instance of running the simulation, I have also calculated the number of plays by adding up the rounds played until the simulation stopped, i.e. when one of the stopping rules, described above in methods, is met. I calculated the average earnings and the average number of plays by taking the mean earnings and mean number of plays from each of the 10,000 rounds of the simulation. 

```{r, cache=TRUE}
# Simulation 1
results <- data.frame(matrix(NA, nrow = 10000, ncol = 2))
results <- rename(results, c("walk_out_money"="X1", "end_num_plays"="X2"))
for(j in seq_along(results$walk_out_money)) {
  results[j,] <- one_series(B = 200, W = 300, L = 1000, M = 100) %>% get_last
  walk_out_money <- results$walk_out_money
  end_num_plays <- results$end_num_plays
}

# Walk out money distribution
hist(walk_out_money, breaks = 100)
```

```{r, include=FALSE}
# Estimated probability of walking out with extra cash
mean(walk_out_money > 200)

# Estimated earnings
mean(walk_out_money - 200)

#Average number of plays before stopping
mean(end_num_plays)
```

The above histogram shows that about half the time the gambler would come out with earnings of about $300 after a series of plays and about half the time would come out with 0 or negative earnings aka at a loss. Digging a little further I found the estimated probability of walking out with a profit is `r mean(walk_out_money > 200)` with the estimated average earnings of `r mean(walk_out_money - 200)`. In addition, the average number of times the gambler was able to play before stopping was `r mean(end_num_plays)`. This means that just slightly over half of the time the gambler would make a profit, but over a long period of gambling with this strategy he/she would actually lose $`r abs(round(mean(walk_out_money - 200),2))` on average after playing about `r round(mean(end_num_plays),0)` rounds of Roulette.

To do some analysis on how changing parameters would impact the Martingale strategy simulation, I used the same inputs as above except for the budget threshold which I increased to $500 and ran the simulation another 10,000 times. Let's see how this affected the outcome:

```{r, cache = TRUE}
# Simulation 2 changing one parameter (W - the budget threshold for successfully stopping changed to 500)
results2 <- data.frame(matrix(NA, nrow = 10000, ncol = 2))
results2 <- rename(results2, c("walk_out_money"="X1", "end_num_plays"="X2"))
for(j in seq_along(results2$walk_out_money)) {
  results2[j,] <- one_series(B = 200, W = 500, L = 1000, M = 100) %>% get_last
  walk_out_money2 <- results2$walk_out_money
  end_num_plays2 <- results2$end_num_plays
}

# Walk out money distribution
hist(walk_out_money2, breaks = 100)
```

```{r, include=FALSE}
# Estimated probability of walking out with extra cash
mean(walk_out_money2 > 200)

# Estimated earnings
mean(walk_out_money2 - 200)

#Average number of plays before stopping
mean(end_num_plays2)
```

The above histogram shows that about 20% the time the gambler would come out with earnings of about $500 after a series of plays and about 80% the time would come out with 0 or negative earnings aka at a loss. In this scenario, I found the estimated probability of walking out with a profit is `r mean(walk_out_money2 > 200)` with the estimated average earnings of `r mean(walk_out_money2 - 200)`. Meaning the gambler has a better chance of making more earnings than in the last scenario but at much less frequency. Over a long period of gambling with this strategy he/she would actually lose $`r abs(round(mean(walk_out_money2 - 200),2))` on average making this situation more risky. Furthermore, it took the gambler about `r round(mean(end_num_plays2),0)` plays before stopping which is a large increase from the last scenario.

# Conclusion

Although the Martingale strategy sounded like a promising way to defeat the house and have a definite outcome of profits, the simulation shows that this does not hold true in practicality. In this experiment, although the gambler did have a chance to make a good profit in certain rounds of betting using the Martingale strategy, their chances are much lower than anticipated. In fact, on average 
the gambler would actually be at a loss. 

Additionally, through increasing the budget threshold, I found the gambler's chances of making a profit decreased significantly. We do not know how changing any of the others parameters will impact the gambler's chances, but this shows that there are several outside factors that could affect how the Martingale's strategy plays out in a Roulette game. In this scenario, we are limited to the parameters and assumptions made which of course would be different across all casinos. I had to make assumptions about the budget of the gambler, the casino's wager limits, the budget thresholds and the maximum number of plays allowed. As shown, just changing one of these assumptions completely changed the average earnings of the gambler and largely increased the number of plays made before stopping. There are many other factors that could impact the outcome of the Martingale strategy while playing Roulette including which version of Roulette is being played. For example, the European version only includes one green pocket meaning the probability of randomly selecting a red pocket would be 18/37 instead of the 18/38 used in this scenario. 

In conclusion, we know that there are many additional factors and assumptions outside of the Martingale strategy at play that impact a gambler's earnings. Under the assumptions I made in this experiment, the gambler would in fact be at a loss on average playing Roulette using the Martingale strategy, hence this strategy is not an effective strategy of gambling while playing Roulette. 
